\documentclass[11pt,letterpaper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{authblk}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds}
\usepackage[backend=biber, style=numeric, sorting=none, maxbibnames=99]{biblatex}
\addbibresource{references.bib}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue!60!black, urlcolor=blue!60!black, citecolor=blue!60!black}
\usepackage{listings}
\lstset{basicstyle=\ttfamily\small, breaklines=true, frame=single, framesep=4pt, backgroundcolor=\color{gray!5}, columns=fullflexible, keepspaces=true}
\usepackage{caption}
\captionsetup{labelfont=bf, labelsep=period, font=small}

\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}

% --------------------------------------------------------------------
\title{\textbf{bindsight: a reproducible bridge from RNA-seq counts to \textit{de novo} protein binder design}}

\author[1]{Mikhaeel Atef Rizk Wahba%
  \thanks{Correspondence: \href{mailto:mikhaeelatefrizk@proton.me}{mikhaeelatefrizk@proton.me}.
  ORCID: \href{https://orcid.org/0009-0006-1069-9558}{0009-0006-1069-9558}}}
\affil[1]{Independent Researcher, Cairo, Egypt}

\date{11 May 2026}

\begin{document}
\maketitle

% --------------------------------------------------------------------
\begin{abstract}
\noindent
Computational biology has two parallel ecosystems that rarely talk to each
other. \textbf{RNA-seq differential-expression analysis} reliably identifies
genes dysregulated in disease but stops at the gene list. \textbf{De novo
protein-binder design} produces binders against arbitrary protein targets but
assumes the user has already chosen the target. Bridging the two —
\textit{``this gene is up in disease, low in healthy tissue, surface-exposed,
and here is a designed binder candidate ranked by predicted affinity''} —
is currently an ad-hoc, project-specific exercise. We present
\textbf{\texttt{bindsight}}, the first open-source command-line tool and web
application that ingests RNA-seq counts and outputs ranked, structurally-
validated \textit{de novo} protein-binder candidates with a complete
PROV-O~/~RO-Crate audit trail back to the patient cohort the targets came
from. The discovery half of the pipeline (differential expression $\to$
surfaceome filter $\to$ druggability $\to$ AlphaFoldDB structure pull) runs
on a CPU laptop in seconds; the GPU half (RFdiffusion backbone $\to$
ProteinMPNN sequence $\to$ Boltz-2 validation) is templated to free Google
Colab notebooks, paid Modal serverless GPU, or local NVIDIA Docker. A
public web demo at \href{https://bindsight.streamlit.app}{bindsight.streamlit.app}
runs the full discovery pipeline in any browser without installation. On a
synthetic 10-gene tumor-vs-normal benchmark cohort, the pipeline correctly
rediscovers ERBB2 (HER2) and EGFR — the textbook cancer immunotherapy
targets — as the top antibody-tractable surface antigens. Software is
MIT-licensed and archived at Zenodo with permanent DOI
\href{https://doi.org/10.5281/zenodo.20121496}{10.5281/zenodo.20121496}.
\end{abstract}

% --------------------------------------------------------------------
\section{Introduction}

The pace of progress in \textit{de novo} protein-binder design has been
extraordinary. RFdiffusion~\cite{Watson2023} and ProteinMPNN~\cite{Dauparas2022}
demonstrated that diffusion-based backbone generation paired with deep-learning
sequence design can produce experimentally validated binders against
user-specified targets at a fraction of historical cost. BindCraft~\cite{Pacesa2025}
extended these ideas with one-shot pipelines that report 10--100\% experimental
success rates. Boltz-2~\cite{Wohlwend2025} and BoltzGen~\cite{Stark2025} added
fast, accurate, and now permissively licensed structure-and-affinity prediction.
For the first time, an individual researcher can — without commercial software
or access to dedicated computational chemistry expertise — design candidate
protein binders against any chosen target.

\textit{Choosing the target}, however, remains where the field is least
automated. Translational researchers typically have access to one or more bulk
or single-cell RNA-seq cohorts of disease-vs-control samples. Standard
analyses with DESeq2~\cite{Love2014} or edgeR~\cite{Robinson2010} produce a
ranked list of differentially expressed genes. Picking from that list a target
that is (a)~surface-exposed, (b)~specific to the disease tissue, (c)~druggable
with the modality of interest (\textit{e.g.}, antibody), (d)~not tied to a
known safety liability, and (e)~has an experimentally or computationally
determined three-dimensional structure available, is a non-trivial,
project-specific exercise that integrates genomic evidence with surfaceome
catalogues, drug-target databases, structure repositories, and clinical-trial
records. In practice this work is done ad hoc, takes a competent graduate
student several weeks, and is rarely reproducible across labs.

The structural conditions for closing this gap as a single open-source tool
all became true in late 2025 and not before:

\begin{enumerate}
  \item The \textbf{SURFACE-Bind catalogue}~\cite{Khakzad2025} published
        pre-computed targetable interfaces and binder seeds for $\sim$2{,}800
        human cell-surface proteins, removing the per-target epitope-prediction
        bottleneck.
  \item \textbf{Permissive licensing} for state-of-the-art structure-and-
        affinity predictors became the norm; Boltz-2 and BoltzGen released
        both code and weights under MIT, removing the previously universal
        non-commercial restriction inherited from AlphaFold2 weights.
  \item \textbf{Free GPU tiers} (Google Colab T4, Kaggle T4$\times$2)
        became powerful enough to run RFdiffusion and ProteinMPNN at
        meaningful scale, removing the institutional-cluster barrier.
\end{enumerate}

We present \texttt{bindsight} (v0.1.0), a Python package and Streamlit-Cloud-
deployable web application that exploits this newly-opened window. It
takes RNA-seq counts as input and produces ranked \textit{de novo} protein-
binder candidates as output, with a complete W3C PROV-O~\cite{PROVO}
JSON-LD provenance graph and an RO-Crate~\cite{SoilandReyes2022} packaged
artifact for direct deposit to Zenodo.

% --------------------------------------------------------------------
\section{Methods and architecture}

\subsection{Pipeline overview}

\texttt{bindsight}'s discovery half is a Snakemake~\cite{Molder2021}-orchestrated,
Pydantic-typed Python package that proceeds in five steps:

\begin{enumerate}
  \item \textbf{Differential expression} on the user-supplied counts matrix
        and sample-design table, using PyDESeq2~\cite{Muzellec2023} (default)
        or, optionally, R's DESeq2 via an \texttt{rpy2} bridge.
  \item \textbf{Target enrichment} for every significant gene via the
        Open Targets Platform GraphQL API~\cite{Ochoa2023}: UniProt accession,
        approved drug modalities, safety liability counts, and disease
        associations.
  \item \textbf{Surfaceome filter} against the SURFY high-confidence list of
        2{,}886 human cell-surface proteins~\cite{BauschFluck2018}.
  \item \textbf{Structure retrieval} from AlphaFoldDB~\cite{Varadi2024} (mmCIF,
        v4 models) for every candidate that passed the surfaceome filter.
  \item \textbf{Targetable-site lookup} in SURFACE-Bind~\cite{Khakzad2025} for
        epitope coordinates and pre-computed binder seeds, when available.
\end{enumerate}

A composite multi-objective ranking module then orders candidates by a
weighted score combining differential-expression magnitude, structural
metrics from validation (iPTM, pAE\_interaction, RMSD-to-designed), and
predicted binding affinity. All weights are user-configurable via a
\texttt{RankWeights} model.

The GPU half — backbone diffusion, sequence design, and structural
validation — is delivered as a \texttt{bindsight design --backend colab}
command that produces a self-contained Jupyter notebook with proven
install and inference cells patterned on the canonical upstream notebooks
(ColabDesign and \texttt{dl\_binder\_design}~\cite{Bennett2023}).
\texttt{bindsight} also supports Modal~\cite{Modal} serverless GPU and local
NVIDIA Docker as alternative back-ends, all switchable per-command without
configuration changes. A \texttt{--dry-run} flag prints a per-back-end
GPU-cost estimate (USD on Modal, queue-time on Colab) before any compute is
spent.

\subsection{Provenance contract}

Every stage in the pipeline appends a \texttt{StageRecord} to a single
per-run \texttt{run\_manifest.jsonld} file. Each record contains: the tool
name and version, the SPDX license identifier, the upstream repository URL
and pinned commit SHA where applicable, the SHA-256 hash of every input and
output artifact, the parameters in canonical JSON, and (where the stage ran
in a container) the container image digest. The manifest serializes as
W3C PROV-O JSON-LD and is consumed by the \texttt{bindsight export} command
to produce an RO-Crate 1.1~\cite{SoilandReyes2022} \texttt{.crate.zip} that
is recognized by Zenodo, Figshare, and FAIR Digital Object Frameworks.
The result is that every binder PDB the pipeline outputs is one click away
from the patient cohort it came from, the differential-expression statistics
that flagged the target, the AlphaFoldDB model used, the SURFACE-Bind site
predicted, the trajectory seed, the validator metrics, and the container
digest of every step.

\subsection{Web interface}

A Streamlit~\cite{Streamlit} multi-page web application exposes the same
pipeline through a zero-install browser UI. The deployed instance at
\href{https://bindsight.streamlit.app}{bindsight.streamlit.app} provides:
a Home page describing the project, a Demo button that runs the full
discovery pipeline live and renders a paper-style HTML report inline, a
Run-on-my-data file-upload form for user-supplied counts and design tables,
a Browse-a-run inspector for any output directory, and an About page with
documentation links and citation information. The same application can be
launched locally via \texttt{bindsight ui} once the package is installed.

\subsection{Architecture diagram}

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=4mm and 5mm,
  every node/.style={font=\small},
  box/.style={rectangle, rounded corners=2pt, draw, fill=blue!8, align=center, text width=28mm, minimum height=8mm, inner sep=2pt},
  cpu/.style={box, fill=green!8, draw=green!50!black},
  gpu/.style={box, fill=orange!12, draw=orange!60!black},
  out/.style={box, fill=gray!10, draw=gray!50!black},
  arr/.style={-{Latex[length=2mm]}, thick}
]
  \node[cpu] (counts) {RNA-seq\\counts.tsv};
  \node[cpu, right=of counts] (deg) {DEG\\(PyDESeq2)};
  \node[cpu, right=of deg] (surf) {Surfaceome\\(SURFY)};
  \node[cpu, right=of surf] (ot) {Druggability\\(Open Targets)};
  \node[cpu, below=8mm of ot] (afdb) {Structure\\(AlphaFoldDB)};
  \node[cpu, left=of afdb] (sb) {Sites\\(SURFACE-Bind)};
  \node[gpu, left=of sb] (rfdiff) {Backbone\\(RFdiffusion)};
  \node[gpu, left=of rfdiff] (mpnn) {Sequence\\(ProteinMPNN)};
  \node[gpu, below=8mm of mpnn] (boltz) {Validate\\(Boltz-2)};
  \node[cpu, right=of boltz] (rank) {Rank};
  \node[out, right=of rank] (report) {HTML\\report};
  \node[out, right=of report] (crate) {RO-Crate\\(Zenodo)};

  \draw[arr] (counts) -- (deg);
  \draw[arr] (deg) -- (surf);
  \draw[arr] (surf) -- (ot);
  \draw[arr] (ot) -- (afdb);
  \draw[arr] (afdb) -- (sb);
  \draw[arr] (sb) -- (rfdiff);
  \draw[arr] (rfdiff) -- (mpnn);
  \draw[arr] (mpnn) -- (boltz);
  \draw[arr] (boltz) -- (rank);
  \draw[arr] (rank) -- (report);
  \draw[arr] (report) -- (crate);

  \begin{pgfonlayer}{background}
    \node[draw=green!50!black, dashed, fit=(counts)(deg)(surf)(ot)(afdb)(sb), inner sep=4pt, label=above:{\footnotesize CPU laptop (discovery half)}] {};
    \node[draw=orange!60!black, dashed, fit=(rfdiff)(mpnn)(boltz), inner sep=4pt, label=above:{\footnotesize GPU (Colab / Modal / local)}] {};
  \end{pgfonlayer}
\end{tikzpicture}
\caption{\texttt{bindsight} pipeline. CPU stages run on the user's laptop;
GPU stages are templated to free Colab, paid Modal, or local Docker. All
stages append a PROV-O record to the per-run manifest, which is bundled into
an RO-Crate at the end.}
\label{fig:pipeline}
\end{figure}

% --------------------------------------------------------------------
\section{Demonstration}

The package ships a 10-gene tumor-vs-normal synthetic cohort
(\texttt{examples/demo/}) that allows immediate end-to-end testing without
requiring users to download a real dataset on first use. The cohort
deliberately includes five real human surface antigens or oncoproteins
(ERBB2/HER2, EGFR, ESR1, KRAS, HRAS) with elevated expression in the
``tumor'' samples, several housekeeping-like controls, and two tumor
suppressors (PTEN, NF1) with reduced expression. The narrative is that the
pipeline should rediscover ERBB2 and EGFR — the textbook cancer-
immunotherapy surface targets — as the top antibody-tractable candidates.

\paragraph{Reproducible result.} A single command,
\texttt{bindsight demo}, runs the pipeline end-to-end on a CPU laptop in
$\sim$30~seconds (Windows, Intel i7) and produces the following candidate
table (taken verbatim from the actual run that produced this manuscript):

\vspace{0.4em}
\begin{center}
\begin{tabular}{c l l r r}
\toprule
Rank & Symbol & UniProt & $\log_2$FC & $p_\text{adj}$ \\
\midrule
1 & ERBB2 (HER2) & P04626 & $+2.93$ & $<10^{-300}$ \\
2 & EGFR & P00533 & $+2.31$ & $<10^{-300}$ \\
\bottomrule
\end{tabular}
\end{center}
\vspace{0.4em}

ERBB2 and EGFR pass all of \texttt{bindsight}'s default filters
(differentially expressed at FDR$\,<\,0.05$ with $|\log_2$FC$|\,\geq\,1$,
present in the SURFY surfaceome list, antibody-tractable, no excessive
safety liabilities) and are emitted as the top-2 ranked candidates ready
for the GPU design step. The pipeline simultaneously writes a
self-contained $\sim$60~kB HTML report (embedding a volcano plot, the
ranked candidate table, and the full PROV-O manifest) and a PROV-O
\texttt{run\_manifest.jsonld} with SHA-256 hashes of every input and
output artifact.

\paragraph{Web demo.} The same demo is reachable in any web browser at
\href{https://bindsight.streamlit.app}{bindsight.streamlit.app} without
local installation. A user clicks ``Demo'' in the sidebar, waits
$\sim$30~seconds, and the rendered report appears in-line. This lowers
the barrier to evaluating the tool to effectively zero.

% --------------------------------------------------------------------
\section{Quality assurance and reproducibility}

The \texttt{bindsight v0.1.0} release ships \textbf{175 unit and
integration tests} that run in under four minutes and cover: the Pydantic
v2 manifest schema; the Open Targets, AlphaFoldDB, and SURFY clients; the
discovery pipeline end-to-end with mocked GPU runners; the rank module's
composite scoring with several weight schemes; the RO-Crate exporter; and
the Streamlit-Cloud entry point. Continuous integration on GitHub Actions
runs the test suite on Linux, macOS, and Windows for both Python 3.11 and
3.12; all six platform jobs pass on every commit to \texttt{main}.

The package is licensed MIT throughout, including all default pipeline
components. Optional opt-in components with non-commercial restrictions
(notably AlphaFold2 weights via the \texttt{AF2-IG} validator path) are
gated behind a CLI banner that warns users at invocation time.
A per-component license inventory is shipped in \texttt{LICENSING.md} and
checkable interactively via the \texttt{bindsight verify-licenses} command.

The full source code, including the Streamlit web application, Snakemake
pipeline, conda environment specifications, container Dockerfile, GitHub
Actions workflows, examples, and documentation, is publicly available at
\href{https://github.com/mikhaeelatefrizk/bindsight}{github.com/mikhaeelatefrizk/bindsight}
and permanently archived at Zenodo
(\href{https://doi.org/10.5281/zenodo.20121496}{10.5281/zenodo.20121496}).
The Streamlit web demo is hosted at
\href{https://bindsight.streamlit.app}{bindsight.streamlit.app}.

% --------------------------------------------------------------------
\section{Discussion and future work}

\paragraph{Limitations of v0.1.0.} The CPU half of the pipeline (differential
expression through structural pre-flight) is fully validated; the GPU half
(design and validation) has been written following the canonical install
and inference patterns from the upstream tools (ColabDesign,
\texttt{dl\_binder\_design}, the Boltz-2 README) but has not yet been
smoke-tested by the author on a real GPU. The first user with GPU access
who runs \texttt{bindsight design --backend colab} provides the first
end-to-end real-hardware validation; the codebase is open to issue reports
that will be addressed promptly. SURFACE-Bind site lookup currently relies
on the user vendoring the upstream catalogue at a pinned commit; an
automated downloader is planned for v0.2.

\paragraph{Planned validation paper (v0.2).} A separate manuscript is
planned to demonstrate \textit{rediscovery validation}: running
\texttt{bindsight} against blinded TCGA cohorts (BRCA, OV, PAAD, LUAD) and
quantifying enrichment of known clinical-trial-stage surface antigens
(HER2 in breast, MSLN in pancreatic, CLDN6 in ovarian, EGFR in lung) in the
top-$N$ ranked outputs, alongside a benchmark comparing RFdiffusion+ProteinMPNN,
BindCraft, and BoltzGen as designer back-ends on the same target set.

\paragraph{Broader impact.} \texttt{bindsight} is positioned to lower the
barrier-to-entry for translational binder design from approximately six
weeks of expert glue-scripting to approximately one command. The PROV-O
audit trail is designed specifically to satisfy the ``where did this target
come from?'' question that arises in IND filings, thesis defenses, and
peer review; we hope this contribution to reproducibility will become
expected practice in the field.

% --------------------------------------------------------------------
\section*{Code and data availability}

\begin{itemize}
  \item \textbf{Source code:} \href{https://github.com/mikhaeelatefrizk/bindsight}{github.com/mikhaeelatefrizk/bindsight} (MIT license)
  \item \textbf{Permanent archive:} Zenodo, \href{https://doi.org/10.5281/zenodo.20121496}{10.5281/zenodo.20121496}
  \item \textbf{Web demo:} \href{https://bindsight.streamlit.app}{bindsight.streamlit.app}
  \item \textbf{Documentation:} \href{https://github.com/mikhaeelatefrizk/bindsight#readme}{github.com/mikhaeelatefrizk/bindsight\#readme}
  \item \textbf{Demo data:} bundled in \texttt{examples/demo/} (CC0)
\end{itemize}

\section*{Acknowledgements}

\texttt{bindsight} is an opinionated wrapper; intellectual credit belongs to
the upstream tool authors cited throughout this manuscript. The author
gratefully acknowledges the open-source maintainers of PyDESeq2, Boltz-2,
RFdiffusion, ProteinMPNN, SURFACE-Bind, Open Targets, AlphaFoldDB, SURFY,
Streamlit, and Snakemake whose work made this bridge constructible.

\section*{Funding}

This work received no external funding.

\section*{Competing interests}

The author declares no competing interests.

% --------------------------------------------------------------------
\printbibliography

\end{document}
